Programming Language Foundations in Agda

### 2019-07-18 papers/scp/PLFA.tex
Minor changes
small fixes to SCP paper

@@

@@

@@

The book is broken into three parts. The first part, Logical Foundations,
develops the needed formalisms.  The second part, Programming Language
Foundations, introduces basic methods of operational semantics.
The third part, Denotational Semantics, introduces a simple
model of semantics and its properties.
(SF is divided into books, the first two of which have the same names
as the first two parts of PLFA, and cover similar material.)
Part~I and Part~II up to Untyped were written by Phil,
Part~II from Substitution and Part~III were written by Jeremy.

@@

\paragraph{More: More constructs of simply-typed lambda calculus}
Introduces product, sum, unit, and empty types as well as lists and let bindings
are explained.  Typing and reduction rules are given informally; a few
are then give formally, and the rest are left as exercises for the reader.
The inherently typed representation is used.

@@

SF has a third volume, written by Andrew Appel, on Verified Functional
Algorithms. I'm not sufficiently familiar with that volume to have a view on
whether it would be easy or hard to cover that material in Agda. And SF recently
added a fourth volume on random testing of Coq specifications using QuickChick.
There is currently no tool equivalent to QuickChick available for Agda.

@@

To give a flavour of how the texts compare, we show the
proof of progress for simply-typed lambda calculus from both texts.
Figures~\ref{fig:plfa-progress-1} and \ref{fig:plfa-progress-2}
are taken from PLFA, Chapter Properties,
while Figures~\ref{fig:sf-progress-1} and \ref{fig:sf-progress-2}
are taken from SF, Chapter StlcProp.
Both texts are intended to be read online,
and the figures were taken by grabbing bitmaps of the text as
displayed in a browser.

@@

In Agda, interaction consists of writing code with holes, at each
point examining the current goal and the variables in scope, and
typing code or executing an Emacs command.  The number of commands
available is much smaller than with Coq, the most important ones being
to show the type of the hole and the types of the variables in scope;
to check the code; to do a case analysis on a given variable; or to
guess how to fill in the hole with constructors or variables in scope.
An Agda proof consists of typed code.  The interaction is \emph{not}
recorded.  Students may recreate it by commenting out bits of code and
introducing a hole in their place.   PLFA contains some prose descriptions
of interactively building code, but mainly contains code that students
can read.  They may also introduce holes to interact with the code, but
we expect this will be rarer than with SF.

@@

% NEW
Some researchers are clearly familiar with the connection between
progress and preservation and animation.  In private correspondence,
Bob Harper referred to it as the \emph{pas de deux}, a dance between
progress, which takes well-typing to a step, and preservation, which
takes a step back to well-typing.  Nonetheless, neither the technique
nor the appealing terminology, appears in \citet{Harper-2016}.  The
appeal to the Agda mailing list bore late fruit: Oleg Kiselyov
directed me to unpublished remarks on his web page where he uses the
name \texttt{eval} for a proof of progress and notes ``the very proof
of type soundness can be used to evaluate sample expressions''
\citep{Kiselyov-2009}.  Nonetheless, as of this writing, we still have
not located a mention in the published literature.

% NEW
There are places in the literature where one might expect a remark on
the relation between progress and preservation and animation---but no
such remark appears.  In the PoplMark Challenge
\citep{Ayedemir-et-al-2005}, Challenge~2A is to prove progress and
preservation for System F$_{<:}$, while Challenge~3 is to prove
animation for the same system. Nowhere do the authors indicate that in
an intuitionistic logic these are essentially the same problem.
\cite{Owens-et-al-2016}, when discussing extraction of animators for
small-step semantics, mention Redex and K, but no other possibilities.
We hope the stress in PLFA on the fact that in an intuitionistic
setting progress and preservation imply animation will mean that the
connection becomes more widely known.

@@

Philip now has five years of experience teaching from SF and one year
teaching from PLFA.  To date, he has taught three courses from PLFA.
\begin{itemize}
\item
  University of Edinburgh, September--December 2018 (with teaching
  assistance from Wen and Chad Nester); twenty two-hour slots,
  comprising one hour of lecture followed by one hour of lab. Ten
  students completed the course, fourth-year undergraduates and
  masters.  The course covered Parts~I and~II of PLFA.

\item
  Pontifícia Universidade Católica do Rio de Janeiro (PUC-Rio),
  March--July 2019, hosted by Roberto Ieuramalischy; ten three-hour
  slots, comprising two hours of lecture followed by one hour of lab.
  Ten students completed the course, mostly doctoral students.  The
  course covered Parts~I and~II of PLFA, save students read chapter
  Lists on their own, and chapter Bisimilarity was skipped.

\item
  University of Padova, June 2018, hosted by Maria Emilia Maietti; two
  three-hour slots, comprising two hours of lecture followed by one
  hour of lab.  Thirty undergraduate students sat the course, which
  covered chapters Naturals, Induction, and Relations.

@@

Students must answer question 1 on the exam, and one of questions 2 or
3.  Question 1 gives predicates over a data structure, such lists or
trees, to be formalised in Agda, and a theorem relating the
predicates, to be formalised and proved in Agda.  Question 2 gives the
students the intrinsic formulation of lambda calculus from chapter
DeBruijn, which they must extend with a described language feature.
Question 3 give the students the the bidirectional type inferencer
from chapter Inference, which they must extend with a described
language feature.

Because the course is taught using a proof assistant, it is important
that students have access to a proof assistant during the exam.
Students are told in advance that they are expected to get perfect on
the exam, and that they will have to study hard to achieve this level.
Given that the goal of formal methods is to avoid error, we believe a
pedagogical purpose is served by telling the students that they are
expected to achieve perfection and making it possible for them to do
so.  Students are given two opportunities to practice in the run up to
the exam, a `mock' exam given in class under exam conditions (two
hours online), and before that a `mock mock' exam as coursework (in
their own time, encouraged to ask questions, tasked to do all three
questions rather than two of three).

For the courses run at Edinburgh and PUC-Rio, the scores vary
widely on the mock, but all students achieve perfection on the
exam.  (The one exception was a PUC-Rio student who did not attend
classes or sit the mock.)  Similar results were achieved at Edinburgh
over the previous five years, using SF as the course textbook and Coq
as the proof assistant.  We consider these results a tribute to the
students' ability to study and learn.

@@

The book is built, tested, and published after each commit, using Travis CI, a
web service for continuous integration.  The means that the book is constantly
changing.  To accommodate those who want a more stable version, e.g., for
teaching, we maintain a stable version of the book at
\begin{center}
  \url{http://plfa.inf.ed.ac.uk/stable}
\end{center}
The stable version of the book is only updated much less frequently, and updates
are announced.

We maintain a tool called, simply, \texttt{acknowledgements}, which uses
the GitHub API to automatically extract a list of contributors to the book, and
add them to the Acknowledgements page, each time the book is published.  We
consider anyone who has sent a successful pull requests a contributor, and sort
contributors in the acknowledgments by the number of accepted requests.
Arguably, a different metric, such as total number of affected lines, might be
more appropriate, though any solution may have its flaws.

@@

There is much left to do!  We hope others may be inspired to join us and expand
the book.

>>>>

% WRITE TO JEREMY RE: DIFFERENCE IN STYLE OF CHAPTER TITLES
% NEED TO CITE SBMF PAPER AND DESCRIBE DIFFERENCES
% STRIP OUT EXAM RESULTS
% CHANGE INHERENT --> INTRINSIC, CITE REYNOLDS

@@

\hyphenation{coq-IDE}

@@

This paper is the journal version of \citet{Wadler-2018}.
It adds two new authors, summaries of Jeremy's new chapters,
and sections on experience with teaching and software
used to publish the book.  The original text often used
first person, which here is replaced by reference to Philip.

@@

The book is broken into three parts. The first part, Logical Foundations,
develops the needed formalisms.  The second part, Programming Language
Foundations, introduces basic methods of operational semantics.
The third part, Denotational Semantics, introduces a simple
model of semantics and its properties.
(SF is divided into books, the first two of which have the same names
as the first two parts of PLFA, and cover similar material.)
Part~I and Part~II up to Untyped were written by Philip,
Part~II from Substitution and Part~III were written by Jeremy.

@@

\paragraph{More: More constructs of simply-typed lambda calculus}
Introduces product, sum, unit, and empty types; and explains lists and let bindings.
Typing and reduction rules are given informally; a few
are then give formally, and the rest are left as exercises for the reader.
The inherently typed representation is used.

@@

SF has a third volume, written by Andrew Appel, on Verified Functional
Algorithms. We are not sufficiently familiar with that volume to have a view on
whether it would be easy or hard to cover that material in Agda. And SF recently
added a fourth volume on random testing of Coq specifications using QuickChick.
There is currently no tool equivalent to QuickChick available for Agda.

@@

To give a flavour of how the texts compare, we show the
proof of progress for simply-typed lambda calculus from both texts.
Figures~\ref{fig:plfa-progress-1} and \ref{fig:plfa-progress-2}
are taken from PLFA, Chapter Properties,
while Figures~\ref{fig:sf-progress-1} and \ref{fig:sf-progress-2}
are taken from SF, Chapter StlcProp.
Both texts are intended to be read online,
and the figures show screengrabs of the text as
displayed in a browser.

@@

In Agda, interaction consists of writing code with holes, at each
point examining the current goal and the variables in scope, and
typing code or executing an Emacs command.  The number of commands
available is much smaller than with Coq, the most important ones being
to show the type of the hole and the types of the variables in scope;
to typecheck the code; to do a case analysis on a given variable; or to
search for a way to fill in the hole with constructors or variables in scope.
An Agda proof consists of typed code.  The interaction is \emph{not}
recorded.  Students may recreate it by commenting out bits of code and
introducing a hole in their place.   PLFA contains some prose descriptions
of interactively building code, but mainly contains code that students
can read.  They may also introduce holes to interact with the code, but
we expect this will be rare.

@@

% NEW
Some researchers are clearly familiar with the connection between
progress and preservation and animation.  In private correspondence,
Bob Harper referred to it as the \emph{pas de deux}, a dance between
progress, which takes well-typing to a step, and preservation, which
takes a step back to well-typing.  Nonetheless, neither the technique
nor the appealing terminology appears in \citet{Harper-2016}.  The
appeal to the Agda mailing list bore late fruit: Oleg Kiselyov
directed me to unpublished remarks on his web page where he uses the
name \texttt{eval} for a proof of progress and notes ``the very proof
of type soundness can be used to evaluate sample expressions''
\citep{Kiselyov-2009}.  Nonetheless, as of this writing, we still have
not located a mention in the published literature.

% NEW
There are places in the literature where one might expect a remark on
the relation between progress and preservation and animation---but no
such remark appears.  In the PoplMark Challenge
\citep{Aydemir-et-al-2005}, Challenge~2A is to prove progress and
preservation for System F$_{<:}$, while Challenge~3 is to prove
animation for the same system. Nowhere do the authors indicate that in
an intuitionistic logic these are essentially the same problem.
\cite{Owens-et-al-2016}, when discussing extraction of animators for
small-step semantics, mention Redex and K, but no other possibilities.
We hope the stress in PLFA on the fact that in an intuitionistic
setting progress and preservation imply animation will mean that the
connection becomes more widely known.

@@

Philip now has five years of experience teaching from SF and one year
teaching from PLFA.  To date, he has taught three courses from PLFA.
\begin{itemize}
\item
  University of Edinburgh, September--December 2018 (with teaching
  assistance from Wen and Chad Nester); twenty 2-hour slots,
  comprising one hour of lecture followed by one hour of lab. Ten
  students completed the course, fourth-year undergraduates and
  masters.  The course covered Parts~I and~II of PLFA.

\item
  Pontifícia Universidade Católica do Rio de Janeiro (PUC-Rio),
  March--July 2019, hosted by Roberto Ieuramalischy; ten 3-hour
  slots, comprising two hours of lecture followed by one hour of lab.
  Ten students completed the course, mostly doctoral students.  The
  course covered Parts~I and~II of PLFA, save students read chapter
  Lists on their own, and chapter Bisimilarity was skipped.

\item
  University of Padova, June 2018, hosted by Maria Emilia Maietti; two
  3-hour slots, comprising two hours of lecture followed by one
  hour of lab.  Thirty undergraduate students sat the course, which
  covered chapters Naturals, Induction, and Relations.

@@

Students must answer question~1 on the exam, and one of questions~2
or~3.  Question~1 gives predicates over a data structure, such lists or
trees, to be formalised in Agda, and a theorem relating the
predicates, to be formalised and proved in Agda.  Question~2 gives the
students the inherent formulation of lambda calculus from chapter
DeBruijn, which they must extend with a described language feature.
Question~3 give the students the the bidirectional type inferencer
from chapter Inference, which they must extend with a described
language feature.

Because the course is taught using a proof assistant, it is important
that students have access to a proof assistant during the exam.
Students are told in advance that they are expected to get perfect on
the exam, and that they will have to study hard to achieve this level.
Given that the goal of formal methods is to avoid error, we believe a
pedagogical purpose is served by telling the students that they are
expected to achieve perfection and making it possible for them to do
so.  Students are given two opportunities to practice in the run up to
the exam, a ``mock'' exam given in class under exam conditions (two
hours online), and before that a ``mock mock'' exam as coursework (in
their own time, encouraged to ask questions, tasked to do all three
questions rather than two of three).

% Table~\ref{tab:exams} shows performance on the mock exam and the final
% exam for the courses run at Edinburgh and PUC-Rio.  While scores vary
% widely on the mock, all students achieve perfection on the
For the courses run at Edinburgh and PUC-Rio, the scores vary
widely on the mock: minimum <20%, maximum 100%, mean
77.8, standard deviation 27.6. But all students achieve perfection on the
exam.  (The one exception was a PUC-Rio student who did not attend
classes or sit the mock.)  Similar results were achieved at Edinburgh
over the previous five years, using SF as the course textbook and Coq
as the proof assistant.  We consider these results a tribute to the
students' ability to study and learn.

% \begin{table}
% \begin{center}
%   \begin{tabular}{|cc|cc|}
%     \hline
%     \multicolumn{2}{|c|}{Edinburgh} &
%     \multicolumn{2}{c|}{PUC-Rio} \\
%     \hline
%     Mock & Exam & Mock & Exam \\
%     \hline
%      15  &  50  &  --  &  40  \\
%      29  &  50  &  ~6  &  50  \\
%      33  &  50  &  20  &  50  \\
%      35  &  50  &  28  &  50  \\
%      36  &  50  &  41  &  50  \\
%      48  &  50  &  49  &  50  \\
%      49  &  50  &  50  &  50  \\
%      50  &  50  &  50  &  50  \\
%      50  &  50  &  50  &  50  \\
%      50  &  50  &  50  &  50  \\
%     \hline
%   \end{tabular}
% \end{center}
% \caption{Exam marks}
% \label{tab:exams}
% \end{table}

@@

The book is built, tested, and published after each commit, using Travis CI, a
web service for continuous integration.  This means that the book is constantly
changing.  To accommodate those who want a more stable version, e.g., for
teaching, we maintain a stable version of the book at
\begin{center}
  \url{http://plfa.inf.ed.ac.uk/stable}
\end{center}
The stable version of the book is updated much less frequently, and updates
are announced.

We maintain a tool called, simply, \texttt{acknowledgements}, which uses
the GitHub API to automatically extract a list of contributors to the book, and
add them to the Acknowledgements page, each time the book is published.  We
consider anyone who has sent a successful pull request a contributor, and sort
contributors in the acknowledgments by the number of accepted requests.
Arguably, a different metric, such as total number of affected lines, might be
more appropriate, though any solution will have its flaws.

@@

There is much left to do!  We hope others may be inspired to join us
to expand and improve the book.
